Wait-free Queues with Polylogarithmic Step Complexity

### 2022-09-21 thesis/thesis.tex
review pages 21,22

%\Statex \bf{Structure}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
\end{itemize}

@@

%\Statex $\diamondsuit$ \tt{\sl{Structures}}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
%\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
\end{itemize}

%\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
%\begin{itemize}
%  
%\item \tt{\sl{PBRT} blocks}
%
%  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
%  
%\end{itemize}

%\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
%\begin{itemize}

%\end{itemize}

%\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
%\begin{itemize}
%  
%  \item \tt{\sl{int} \tt{last\sub{done}}}
%
%  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the. A block is finished if all of its operations are finished. \tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response. \it{put the definitions before the pseudocode}}
%  
%\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Block}} 
%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}

\begin{itemize}
%  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
%  \textsf{: \# enqueue, dequeue operations in the block}
%  \item \tt{\sl{int} num, sum}
%  \textsf{: total \# operations in block, prefix sum of \tt{num}}

  \item \tt{\sl{int} super}
  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} \size}
  \textsf{: size of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
%  \item \tt{\sl{counter} num\sub{finished}}
%  \textsf{: number of finished operations in the block}
%    \item \tt{\sl{int} order}
%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

@@

%\it{Variable naming:}
%\begin{itemize}
%  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
%  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
%\end{itemize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
%\If{\tt{i\sub{enq}==-1}}
%\State \tt{output= null}
%\Else
%\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
%\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\begin{algorithm}
\caption{BlockList}
\begin{algorithmic}[1]
\setcounter{ALG@line}{700}

\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}. Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}. If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}

%\Statex
%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
%\State \tt{l.Help()}
%\EndFor
%\EndIf
%\State \tt{blk.num\sub{finished}= 0}
%%\State \tt{*oldRoot= \&root.blocks.root}
%%\State \tt{*newRoot= root.blocks.Append(blk).root}
%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
%\State \Return{\tt{CAS(blocks[n], null, blk)}}
%\EndFunction{TryAppend}

\Statex
%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
\Statex
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

\end{algorithmic}
\end{algorithm}
%##########################################

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\State \tt{hp= parent.\head} \label{readParentHead1}
%\If{\nf{blocks[h]!=null}}
%\State \tt{CAS(blocks[h].super, null, hp)} \label{helpSuper1}
%\EndIf
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}}
\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
%\cmt{Created block contains nothing.}
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}

\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
%\cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{result}}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchEnd}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

%\begin{algorithm}
%\caption{Root}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{500}
%
%\Function{Block}{FindMostRecentDone}{}
%\For{\tt{leaf l} \bf{in leaves}}
%\State\tt{max= Max(l.maxOld, max)}
%\EndFor
%\State\Return\tt{max} \cmt{This snapshot suffies.}
%\EndFunction{FindMostRecentDone}
%
%\end{algorithmic}
%\end{algorithm}
%##########################################

%\Statex
%
%\Function{void}{Help}{}\cmt{Helps pending operations}
%
%\State{\tt{last= l.\head-1}}
%\cmt{\tt{l.blocks[last]} can not be \tt{null} because \head increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
%\EndIf
%
%\EndFunction{Help}

%\begin{algorithm}
%\caption{Yet to decide how to handle.}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{800}
%
%%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
%%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
%%\State \tt{t1,t2= RBT.split(order, s)}
%%\State \tt{RBTRoot.CAS(t2.root)}
%%\EndFunction{CollectGarbage}
%
%\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
%
%\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
%
%\State \Return \tt{this.leaf.response(block.order)}
%\EndIf
%
%\EndFunction{FallBack}
%
%\end{algorithmic}
%\end{algorithm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                     %
                     %
                     %
                     %
                     %
                     %
                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To prove a shared data structure works correctly it is sufficient to show it is linearizable. In our case which we create the linearization ordering in the root, we need to prove the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and if we do the linearization ordering created on a sequential queue we get the same results as our queue. The proof is structured like this. First we define and prove some facts about blocks and node's head properties. Then we introduce  the linearization ordering formally. Then we prove double \nf{Refresh} on a node is enough to propagate children's new operations up to the node. After this we prove some claims about the size of each block. Having propagates working  in a wait-free manner then we prove the correctness of \nf{DSearch()} and \nf{GetENQ()} and \nf{IndexDEQ}. Having these two main methods we prove the way we compute the response of a dequeue is correct and finally we prove the linearizability.

%\framebox[1.1\width]{TEST} Fix the logical order of definitions (cyclic refrences).
%
%%\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof. Is lemma 7 necessary? Is lemma 13 induced trivially from lemma 8?
%
%\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?
%
%%\framebox[1.1\width]{TEST} How to merge notions of blocks and operations? block b $\sqsubseteq$ block c means b is subblock of c. block b $\in$ set B means b is in B. Merge these two to have shorter formulaes.
%
%\framebox[1.1\width]{Question} A good notation for \it{the index of the \nf{b}}?
%
%\framebox[1.1\width]{Question} How to remove the notion of time? To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t. Is it good? Furthermore, can we remove the notion of established blocks?

A block is an object storing some statistics, as described in Algorithm Queue. A block in a node's blocklist implicitly represents a set of operations.
\begin{definition}[Ordering of a block in a node]
Let $b$ be \nf{n.blocks[$i$]} and $b^\prime$ be \nf{n.blocks[$j$]}. We call $i$ the \emph{index} of block $b$. Block $b$ is \it{before} block $b^\prime$ in node \nf{n} if and only if $i<j$. We define \it{the prefix for} $b$, to be the blocks in the \tt{n.blocks[0..$i$]}.
\end{definition}

Next we show that \nf{head} in a node can only be increased and when a \nf{Refresh} terminates \nf{head} is incremented by the process itself or maybe helped by another process.
\begin{observation} \label{nonDecreasingHead}
  \nf{head} is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()} .
\end{proof}
\begin{lemma}[head Increment] \label{lem::headInc}
Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
\end{lemma}
\begin{proof}
If the \nf{CAS} in Line \ref{incrementHead} is successful then the claim holds. Otherwise it has changed from the value that was read in Line \ref{readHead}, which by Observation \ref{nonDecreasingHead} it means another process has incremented the head.
\end{proof}

@@

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is $h$ then $\nf{n.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{n.blocks[i]}\neq\nf{null}$ for ${i<h}$.
\end{invariant}
\begin{proof}
The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[}$x$\nf{]} is null for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show the invariant still holds after these two changes.

In the algorithm, \nf{n.blocks} is modified only on Line \ref{cas} which updates \nf{n.blocks[$h$]} for $h$ read from \nf{n.head} in Line \ref{readHead}.Note that since the \nf{CAS} in Line \ref{cas} is successful it means \nf{n.head} has not changed from $h$ till doing the \nf{CAS}. Writing into \nf{n.blocks[n.head]} preserves the invariant, since the claim does not talk about the content of \nf{n.blocks[n.head]}.

The value of \nf{n.head} is modified only in Line \ref{incrementHead}. If \nf{n.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. When a process does Line \ref{incrementHead} it means Line \ref{cas} was finished which either case it is successful or not it means \nf{n.blocks[$h$]} is not \nf{null} anymore. %At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}

The next lemma is used to prove the subblocks of two blocks in a node are disjoint. (I can't say its use without mentioning subblocks)
\begin{lemma}[headProgress] \label{lem::headProgress}
 If $\nf{n.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}

Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}. $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked. The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was minus one of the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}. Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was minus one of the value read from $\nf{n.left.head}-1$ during the call to \nf{CreateBlock($i-1$)}. As by Observation \ref{nonDecreasingHead} \nf{n.left.head} is non-decreasing, $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}

We define the subblocks of a block recursively.
\begin{definition}[Subblock]\label{def::subblock}
A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]} . Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.  We say block $b$ is \it{propagated to node} node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n.blocks$.
\end{definition}

>>>>

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
\item \tt{\sl{blocks}} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Block}}

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} \size}
  \textsf{: size of the queue after performing all operations in the prefix for this block}
\end{itemize}

@@

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchEnd}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}

\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
\State \Return{ \tt{result}}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
\EndIf

@@

@@

@@

To prove a shared data structure works correctly, it is sufficient to show it is linearizable. In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization operations get the same results as in our queue. The proof is structured like this. First, we define and prove some facts about blocks and the node's \nf{head} field. Then, we introduce  the linearization ordering formally. Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1). After, this we prove some claims about the size and operation of each block which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}. Finally (2) is followed by proving the correctness of the way we compute the response of a dequeue.

A block is an object storing some statistics, as described in Algorithm Queue. A block in a node's blocklist implicitly represents a set of operations.
\begin{definition}[Ordering of a block in a node]
Let $b$ be \nf{$n$.blocks[$i$]} and $b^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $b$. Block $b$ is \it{before} block $b^\prime$ in node \nf{n} if and only if $i<j$. We define \it{the prefix for} $b$ to be the blocks in \tt{n.blocks[0..$i$]}.
\end{definition}

Next, we show that the value of \nf{head} in a node can be increased. After the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
\begin{observation} \label{nonDecreasingHead}
For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()} .
\end{proof}

\begin{lemma} \label{lem::headInc}
Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates, \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
\end{lemma}
\begin{proof}
If the \nf{CAS} in Line \ref{incrementHead} is successful then the claim holds. Otherwise $n.\nf{head}$ has changed from the value that was read in Line \ref{readHead}. By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
\end{proof}

@@

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
\end{invariant}
\begin{proof}
Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}\neq\nf{null}$ for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show the invariant still holds after these two changes.

In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}. Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}, because if so then $n.\nf{blocks[$h$].CAS}$ could not be successful. Writing into \nf{$n$.blocks[$n$.head$=h$]} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.

The value of \nf{n.head} is modified only in Line \ref{incrementHead}. If \nf{n.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. \nf{Advance()} is called  in Lines \ref{advance} and \ref{helpAdvance}. In case \nf{advance}, Line \ref{cas} was finished  before doing \ref{incrementHead} wether \ref{cas} is successful or not $n.\nf{blocks[h]}\neq\nf{null}$. In the other case also $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
\end{proof}

We define the subblocks of a block recursively.

\begin{definition}[Subblock]\label{def::subblock}
A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]}. Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.  We say block $b$ is \it{propagated} to node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n\nf{.blocks}$.
\end{definition}

The next lemma is used to prove the subblocks of two blocks in a node are disjoint.
\begin{lemma} \label{lem::headProgress}
 If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}
Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}. Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked. The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}. Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}. By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}
